---
title: "eds223-homework-2"
format: html
editor: visual
---

# Redlining and environmental inustice:

## Peter Vitale

In this project tktktkt:

-   

-   

I'm going to start by reading in the data on ()

```{r}
 #| message: false 
 #| warning: false
 #First, I'm going to load the necessary packages using p_load 
pacman::p_load('tidyverse', 
               'sf', 
               'here', 
               'tmap', 
               'kableExtra',
               'patchwork',
               'ggthemes')
```

### Data Aggregation
Read in geodatabase of EJScreen data at the Census Block Group level

```{r}
ejscreen <- sf::st_read(here::here("data", "ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

#Lets filter that to be a bit more specific 

california <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "CA") 

# Filter to San Francisco Bay 
la_county <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("Los Angeles County", "Orange County", "San Bernardino County")) %>%   # These counties make up our area
  st_transform(crs = st_crs(inequality)) %>%  # Making the CRS match the other datasets
  filter(AREALAND != 0) # Turns off the coast, making plotting prettier

gbif <-  sf::st_read(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))

inequality <-  sf::st_read(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))
```

## Part One

### 1.1 
I am now going to make a map of the HOLC redlining.

```{r}
tm_shape(la_county, bbox = st_bbox(inequality))+ # We set the bbox as inequality
  tm_crs(4326)+
  tm_polygons(fill = 'grey',lty = 'blank')+ # Fill for my base map
  tm_title('Historic Redlining indicators in Los Angeles County')+
  tm_layout(bg.color = 'blue')+ # Fill for the Water
tm_shape(inequality)+
  tm_polygons(fill = 'grade', 
              fill.scale = tm_scale(values = c('#646D4E','#C1A863', '#879595','#9E5D67' )),
              fill.legend = tm_legend(title = 'Grade', na.show = FALSE))+ 
  # Took na's out of legend
tm_compass(type = 'rose', position = c('right','top'), size = 3)+ # Compass Rose
tm_scalebar(position = c('right','top'))+
  tmap_options(component.autoscale = FALSE) # Turn off autoscale
# I set component.autoscale to false so i could be more in control of the components


```

### 1.2

I want to make a table that summarizes the percentage of census block groups that fall in each grade

I first need to join `la_county` and `inequality`

```{r}

inequality <- st_make_valid(inequality)

la_holc_ej_join <- st_join(inequality, la_county) # Join my two 

la_holc_ej_join <- st_drop_geometry(la_holc_ej_join) # Drop the geometry

# Now I'm going to take a proportion of the census block groups that fall in to each grade. 

percentage_grade_df <- la_holc_ej_join %>% 
  group_by(grade) %>% 
  summarise(num_of = n()) %>% # Find sum of counts
  mutate(grade_percentage = round(x= (num_of / sum(num_of)) * 100, digits = 2)) %>% 
  # This takes the average of the sum
  ungroup()

# And now I can make that table

kbl(percentage_grade_df, 
      col.names = c("Grade", "Count", "Percentage (%)") ) %>% 
    kable_classic(full_width = F, html_font = "Cambria") %>% # Turn off full width
  # Table, use Cambria to make it look like its out of a paper
  footnote(number = "Table 1. Census Block Group Percentage by HOLC Grade")



```

### 1.3

I want to make plots to see how redlining is related to some environmental justice issues

First I need to take means of my variables of interest

```{r}
ej_mean_data <- la_holc_ej_join %>% 
  group_by(grade) %>% 
  summarise(mean_low_income = mean(LOWINCPCT), mean_particulate_percent = mean(P_D2_PM25), mean_low_life_percent = mean(P_LIFEEXPPCT, na.rm = TRUE))
# Here I used the summarise function to get a bunch of means at once. 
```

```{r}
 p1 <- ggplot(ej_mean_data, aes(x = grade, y = mean_low_income, fill = grade))+
  geom_col()+ # This makes a bar graph with x and y axes
  scale_fill_manual(values = c('#ABC798','#C1DABD', '#E6983F','#A0185A' ))+
  labs(x = 'HOLC Grade',
       y = 'Mean Percent Low income')+
  theme_clean()+
  theme(legend.position = 'none')

 p2 <- ggplot(ej_mean_data, aes(x = grade, y = mean_particulate_percent, fill = grade))+
  geom_col()+
  scale_fill_manual(values = c('#ABC798','#C1DABD', '#E6983F','#A0185A' ))+
  labs(x = 'HOLC Grade',
       y = 'Mean Particulate Percentile')+
   theme_clean()+
  theme(legend.position = 'none')
 
 p3 <- ggplot(ej_mean_data, aes(x = grade, y = mean_low_life_percent, fill = grade))+
  geom_col()+
  scale_fill_manual(values = c('#ABC798','#C1DABD', '#E6983F','#A0185A' ))+
  labs(x = 'HOLC Grade',
       y = 'Mean Percentile with Low Life Expectancy')+
   theme_clean()+
  theme(legend.position = 'none')


 (p1|p2|p3)+ # Use patchwork to stitch together
   plot_annotation(
  title = 'Mean Percentile with Low Life Expectancy, Particulate Percentile,',
  subtitle = 'and Percent Low Income in Historically Redlined Communities')
   
```

### 1.4

## Part Two

### 2.1

I want to create a plot of percent of bird observations in each HOLC grade

The first thing I need to do is merge `gbif` and `inequality`

```{r}
#| message: false

holc_bird_df <- st_join(inequality, gbif) 

holc_bird_df <- holc_bird_df %>% 
  group_by(grade) %>% 
  summarize(num_of = n()) %>% 
  mutate(grade_percentage = round(x= (num_of / sum(num_of)) * 100, digits = 2)) %>% 
  ungroup()
```

Now I can make a bar plot comparing bird counts and HOLC Grade

```{r}
ggplot(holc_bird_df, aes(x = grade, y = grade_percentage))+
  geom_col(aes(fill = grade))+
  scale_fill_manual(values = c('#ABC798','#C1DABD', '#E6983F','#A0185A' ))+
  labs(x = 'HOLC Grade',
       y = 'Mean Bird Observations',
       title = 'Percent of Bird Observations in HOLC Grades')+
  theme_clean()+
  theme(legend.position = 'none')
  
```
